---
title: "Spur_Impact"
author: "Eric Hedberg"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
load("Planner/appData.Rdata")
data <- qed_data[["between_random"]][which(qed_data[["between_random"]]$school == 306),-c(1,2,7)]
data$ses <- 1-data$ses
colnames(data) <- c("nonwhite", "ses", "math", "science")
set.seed(101)
```

## How I rigged it

The chance of treatment assignment is based on the math scores z-score. 

```{r}
plot(seq(from = -4, to = 4, by = .1), pnorm(seq(from = -4, to = 4, by = .1)), type = "l", ylab = "Cumul. Prob", xlab = "z")
```


and you see that in the code `pnorm(scale(data$math)-1)` below (I add a bit of an offset with the `-1`). Note that I do not build in any "treatment effect," so whatever we see is an "echo" of math. 


```{r}
data$treat <- rbinom(100,1,pnorm(scale(data$math)-1))

summary(lm(science ~ treat + math +ses+nonwhite, data = data))$coef
```

Next, I wrote a quick function to help do this many times

```{r}
spur_impact <- function(data){
  data$treat <- rbinom(100,1,pnorm(scale(data$math)-1))
  coefs <- summary(lm(science ~ treat + math +ses+nonwhite, data = data))$coef
  rawcoefs <- summary(lm(science ~ treat, data = data))$coef
  
  return(cbind(b_c = coefs[2,1], p_c = coefs[2,4],b_uc = rawcoefs[2,1], p_uc = rawcoefs[2,4]))
}
```

So, we can do this a thousand times.

```{r}
results <- data.frame(t(replicate(1000,spur_impact(data), simplify = TRUE)))
```

If we don't control for the math which, the placebo treatment effect is always significant  
```{r}
mean(results$X4 <= .05)
```

But we if we control for it, it can happen, but not a lot

```{r}
mean(results$X2 <= .05)
```

Finally, suppose there was a treatment effect of .5, and it was related to something you didn't measure and sorta correlated with what you have. This new function does this


```{r}
unobs_spur_impact <- function(data){
  other_var <- scale(rnorm(data$math*.15+data$ses*(-.05)+data$nonwhite*(.05))+rnorm(100))
  treat_effect <- runif(100, .25,.75)
  data$treat <- rbinom(100,1,pnorm(other_var))
  data$science <- data$science+treat_effect*data$treat
  coefs <- summary(lm(science ~ treat + math +ses+nonwhite, data = data))$coef
  rawcoefs <- summary(lm(science ~ treat, data = data))$coef
  
  return(cbind(b_c = coefs[2,1], p_c = coefs[2,4],b_uc = rawcoefs[2,1], p_uc = rawcoefs[2,4]))
}
```

So, we can do this a thousand times.

```{r}
results <- data.frame(t(replicate(1000,unobs_spur_impact(data), simplify = TRUE)))
```

If we don't control for the math which, the placebo treatment effect is always significant  
```{r}
mean(results$X4 <= .05)
```

But we if we control for it, it can happen more often

```{r}
mean(results$X2 <= .05)
```
