---
title: "Lab for blocked cluster randomized designs"
output: word_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r, setup, include=FALSE}
load("../Planner/appData.Rdata")
```

# Case Study

## **High School After-School Robotics Program**

Dr. Carlos -- a computer scientist -- and his students have been running
an after-school outreach program focused on robotics at a nearby high
school in Atlanta. This high school predominately serves Black students
-- which are underrepresented in computer science -- and the students
come from both middle-class and low-SES families. In order to be
admitted to the program, which is free, students had to apply and be
recommended by a teacher. Once admitted, students in the program meet
once a week for 2 hours for 16 weeks. During the lessons, they learn to
program and build robots to complete tasks and, at the end of the
program, they take part in a mini 'competition'. The research team has
been excited to see how the students have responded to the program,
including how this has affected their interest in computer science in
general, as well as in studying a STEM related field in college.
Teachers have reported that students in the program are more engaged in
their science classes.

# Plausible Evaluation Design

## Multisite Cluster Randomized Designs  (fixed block effects and random cluster effects)

Suppose researchers wish to evaluate the effectiveness of a similar
program on student science grades in school. Researchers enroll teachers
and their students in the study during the the fall term. Teachers who
are enrolled into the study are randomized within each school into
either the treatment condition (the program) or a control condition.
Spring science scores on a state-mandated standardized test are used as
the outcome. A set of schools are recruited for the study from a variety
of districts through a professional development conference for state
principals and district membership is not considered in the design. For
planning purposes, researchers use a teacher-level (within-school)
intraclass correlation of .05.

The analysis will employ a two-level hierarchical linear model (students
nested within teachers) with a random intercept at the teacher level and
block covariates for schools to compare the averages of the treatment
and control students.

For simplicity, assume perfect compliance with random assignment and a
perfect response rate.

### Question 1

Using the [planner tool](https://steppcenter.shinyapps.io/Planner/),
find the minimum detectable effect for a `Multisite Cluster Randomized Designs  (fixed block effects and random cluster effects)` that
employed 5 schools, in each of which are 2 teachers randomized into
treatment and 2 randomized into control, each with 20 students, when
power is .8 for a test with a significance level of .05.

```{r, include=FALSE}
compute_mdes(V = blockedclustered_fixed_nocovariate_V(m = 5,
                                                        kt = 2, 
                                                      kc = 2,
                                                      n = 20,
                                                      icc = .05), 
             df = blockedclustered_fixed_nocovariate_df(m = 5,
                                                        kt = 2, 
                                                      kc = 2,
                                                      n = 20), 
             beta = 1-.8, 
             alpha = .05 
)
```

### Question 2

A member of the study team claims that many schools only have a two
teachers. Using the [planner
tool](https://steppcenter.shinyapps.io/Planner/), find the minimum
detectable effect for a `Multisite Cluster Randomized Designs  (fixed block effects and random cluster effects)` that employed 5 schools,
in each of which there is 1 teacher randomized into treatment and 1
randomized into control, each with 20 students, when power is .8 for a
test with a significance level of .05.

```{r, include=FALSE}
compute_mdes(V = blockedclustered_fixed_nocovariate_V(m = 5,
                                                        kt = 1, 
                                                      kc = 1,
                                                      n = 20,
                                                      icc = .05), 
             df = blockedclustered_fixed_nocovariate_df(m = 5,
                                                        kt = 1, 
                                                      kc = 1,
                                                      n = 20), 
             beta = 1-.8, 
             alpha = .05 
)

```

### Question 3

The results from Question 2 are different from Question 1. If the
students and teachers are fixed, schools may need to be added. How many
schools need to be added to the design in Question 2 to get a similar
MDES as what you found for Question 1? Before you answer, consider
whether the total sample sizes in Questions 1 and 2, would you need to
increase blocks proportionally?

```{r, include=FALSE}
compute_mdes(V = blockedclustered_fixed_nocovariate_V(m = 8,
                                                        kt = 1, 
                                                      kc = 1,
                                                      n = 20,
                                                      icc = .05), 
             df = blockedclustered_fixed_nocovariate_df(m = 8,
                                                        kt = 1, 
                                                      kc = 1,
                                                      n = 20), 
             beta = 1-.8, 
             alpha = .05 
)
print(paste("with 8 schools"))
```

### Question 4

After signing up the schools, it was apparent that using a 10th grade
math test as a covariate was possible for the analysis. Earlier analyses
of other cohorts show that the 10th grade math test was correlated with
the focal science test at the teacher level at $\rho = .8$ for an
R-square value of $\rho^2 = .64$, and at the student level at
$\rho = .5$ for an R-square value of $\rho^2 = .25$. With this
covariate, would it be possible to still use the sample design in
Question 2 to get the sensitivity of plan in Question 1?

Using the [planner tool](https://steppcenter.shinyapps.io/Planner/),
find the minimum detectable effect for a `Multisite Cluster Randomized Designs  (fixed block effects and random cluster effects)` that employed 5 schools, in each of which there is 1
teacher randomized into treatment and 1 randomized into control, each
with 20 students, and using the covariate, when power is .8 for a test
with a significance level of .05.

```{r, include=FALSE}
compute_mdes(V = blockedclustered_fixed_covariate_V(m = 5,
                                                        kt = 1, 
                                                      kc = 1,
                                                      n = 20,
                                                      icc = .05,
                                                    R2_1 = .25,
                                                    R2_2 = .64
                                                    ), 
             df = blockedclustered_fixed_covariate_df(m = 5,
                                                        kt = 1, 
                                                      kc = 1,
                                                      n = 20,
                                                      q = 1), 
             beta = 1-.8, 
             alpha = .05 
)
```

Is this covariate enough? What other adjustments could be made?

# References

[Barker, B. S., & Ansorge, J. (2007). Robotics as means to increase
achievement scores in an informal learning environment. *Journal of
research on technology in education*, *39*(3),
229-243.](http://web.b.ebscohost.com.turing.library.northwestern.edu/ehost/pdfviewer/pdfviewer?vid=1&sid=71d566ac-b312-45f8-a4d1-00b37f0e57c7%40sessionmgr103)
