---
title: "Lab for individually randomized designs"
output: word_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
load("../Planner/appData.Rdata")
```

# Case Study A

## Museum Science Spatial: A Program for Pre-Schoolers

A STEM Museum in Florida has had a longstanding partnership with two
researchers at a nearby university. Drs. Andrews and Zheng are
interested in how informal experiences in museums may encourage families
to engage with their children in ways that inspire and support STEM
learning at home. This interest has led them to work with the museum
staff to design a series of exhibits that foster interactions (while at
the museum), while also providing materials to take home, with the hope
of continued engagement. Understanding if this program actually improves
these connections is important, given the continued cost of these
take-home materials. This study focuses on a particular exhibit that
focuses on fostering the development of spatial skills through a series
of building activities.

# Plausible Evaluation Design A

## Individual Random Assignment

Suppose researchers wish to evaluate the effectiveness of a similar
program on parent interaction with their children at home, measured
using a well-established validated scale based on questions answered
through an online survey collection system. The design is to recruit
families as they enter the museum into the study, then randomly assign
them to attend the additional exhibits and receive the additional take
home materials (treatment group) or attend the typical exhibits (control
group). After 30 days, all study families take a survey and their
answers are used to compute an outcome score. These scores are then
analyzed using an independent samples $t$ test to compare the averages
of treatment and control families.

For simplicity, assume perfect compliance with random assignment and a
perfect response rate.

### Question A-1

Using the [planner tool](https://steppcenter.shinyapps.io/Planner/),
find the minimum detectable effect for an `Individually Randomized Design` study in which 20 families are
assigned to the treatment group and 20 families are assigned the control
group when power is .8 for a test with a significance level of .05.

```{r, include=FALSE}
compute_mdes(V = srs_fixed_nocovariate_V(20,20),
             df=srs_fixed_nocovariate_df(20,20),
             beta = 1-.8,
             alpha = .05
)

```

### Question A-2

A member of the study team claims that a better study would use a higher
portion of the study families as treatment. Using the [planner
tool](https://steppcenter.shinyapps.io/Planner/), find the minimum
detectable effect for a study in which 30 families are assigned to the
treatment group and 10 families are assigned the control group when
power is .8 for a test with a significance level of .05. Is this design
more sensitive?

```{r, include=FALSE}
compute_mdes(V = srs_fixed_nocovariate_V(30,10),
             df=srs_fixed_nocovariate_df(30,10),
             beta = 1-.8,
             alpha = .05
)
```

### Question A-3

There is some concern that the study may not be as sensitive as the
researchers need it to be. The **first idea** is to use 60 families,
with 30 in treatment and 30 in control, without a covariate. The
**second idea** is to have each of the 40 study members fill out a small
questionnaire that would lead to a covariate correlated with the outcome
at $\rho = 0.5$ ($\rho^2 = 0.25$).

Using the [planner tool](https://steppcenter.shinyapps.io/Planner/),
find the minimum detectable effect for an `Individually Randomized Design` study in which 30 families are
assigned to the treatment group and 30 families are assigned the control
group, without using a covariate, when power is .8 for a test with a
significance level of .05. This is the sensitivity of the **first
idea**.

```{r, include=FALSE}
compute_mdes(V = srs_fixed_nocovariate_V(nt = 30,nc = 30),
             df=srs_fixed_nocovariate_df(nt = 30,nc = 30),
             beta = 1-.8,
             alpha = .05
)
```

Using the [planner tool](https://steppcenter.shinyapps.io/Planner/),
again, find the minimum detectable effect for an `Individually Randomized Design` study in which 20
families are assigned to the treatment group and 20 families are
assigned the control group, with a covariate is employed where
($\rho^2 = 0.25$), when power is .8 for a test with a significance level
of .05. This is the sensitivity of the **second idea**.

```{r, include=FALSE}
compute_mdes(V = srs_fixed_covariate_V(nt = 20, nc = 20, R2_1 = .25, q = 1),
             df=srs_fixed_covariate_df(nt = 20, nc = 20, q = 1),
             beta = 1-.8,
             alpha = .05
)
```

Which idea is more sensitive?

# Case Study B

## **High School After-School Robotics Program**

Dr. Carlos -- a computer scientist -- and his students have been running
an after-school outreach program focused on robotics at a nearby high
school in Atlanta. This high school predominately serves Black students
-- which are underrepresented in computer science -- and the students
come from both middle-class and low-SES families. In order to be
admitted to the program, which is free, students had to apply and be
recommended by a teacher. Once admitted, students in the program meet
once a week for 2 hours for 16 weeks. During the lessons, they learn to
program and build robots to complete tasks and, at the end of the
program, they take part in a mini 'competition'. The research team has
been excited to see how the students have responded to the program,
including how this has affected their interest in computer science in
general, as well as in studying a STEM related field in college.
Teachers have reported that students in the program are more engaged in
their science classes.

# Plausible Evaluation Design B

## Multisite Individually Randomized Design with Fixed Effects

Suppose researchers wish to evaluate the effectiveness of a similar
program on student science grades in school. Researchers enroll students
in the study during the the fall term. Students who are enrolled into
the study are randomized within each school into either the treatment
condition (the program) or a control condition (whatever they planned to
do after school). Spring science scores on a state-mandated standardized
test are used as the outcome. A set of schools are recruited for the
study from a variety of districts through a professional development
conference for state principals and district membership is not
considered in the design. For planning purposes assume block effects
explain about 10 percent of the total variation.

For simplicity, assume perfect compliance with random assignment and a
perfect response rate.

### Question B-1

Using the [planner tool](https://steppcenter.shinyapps.io/Planner/),
find the minimum detectable effect for a `Multisite Individually Randomized Design with Fixed Effects` study that employed 12 schools, in each of which 10 students are
assigned to the treatment group and 30 students are assigned the control
group when power is .8 for a test with a significance level of .05.

```{r, include=FALSE}
compute_mdes(V = dummyblocked_fixed_nocovariate_V(m = 12, 
                                             nt = 10,
                                             nc = 30,
                                             icc = .10), 
             df = dummyblocked_fixed_nocovariate_df(m = 12, 
                                             nt = 10,
                                             nc = 20), 
             beta = 1-.8, 
             alpha = .05 
             )
```

### Question B-2

A member of the study team claims that a better study would use a
similar portion of students in treatment and control condition. Using
the [planner tool](https://steppcenter.shinyapps.io/Planner/), find the
minimum detectable effect for a `Multisite Individually Randomized Design with Fixed Effects` study that employed 12 schools, in each of which 20 students are
assigned to the treatment group and 20 students are assigned the control
group when power is .8 for a test with a significance level of .05. Is
this design more sensitive?

```{r, include=FALSE}
compute_mdes(V = dummyblocked_fixed_nocovariate_V(m = 12, 
                                             nt = 20,
                                             nc = 20,
                                             icc = .10), 
             df = dummyblocked_fixed_nocovariate_df(m = 12, 
                                             nt = 15,
                                             nc = 15), 
             beta = 1-.8, 
             alpha = .05 
             )

```

### Question B-3

After signing up the schools, it was apparent that using a 10th grade
math test as a covariate was possible for the analysis. Earlier analyses
of other cohorts show that the 10th grade math test was correlated with
the focal science test at $\rho = .8$ for an R-square value of
$\rho^2 = .64$. The hope is that having this covariate can reduce the
total number of schools needed from 12 to 10.

Using the [planner tool](https://steppcenter.shinyapps.io/Planner/),
find the minimum detectable effect for a `Multisite Individually Randomized Design with Fixed Effects` study that employed 10 schools, in each of which 20 students are
assigned to the treatment group and 20 students are assigned the control
group for an analysis that also includes the earlier math score, when
power is .8 for a test with a significance level of .05. Is this design
more sensitive?

```{r, include=FALSE}
compute_mdes(V = dummyblocked_fixed_covariate_V(m = 10, 
                                             nt = 20,
                                             nc = 20,
                                             icc = .10,
                                           R2_1 = .64,
                                           q = 1), 
             df = dummyblocked_fixed_covariate_df(m = 10, 
                                             nt = 15,
                                             nc = 15,
                                             q = 1), 
             beta = 1-.8, 
             alpha = .05 
             )
```

# References

[Barker, B. S., & Ansorge, J. (2007). Robotics as means to increase
achievement scores in an informal learning environment. *Journal of
research on technology in education*, *39*(3),
229-243.](http://web.b.ebscohost.com.turing.library.northwestern.edu/ehost/pdfviewer/pdfviewer?vid=1&sid=71d566ac-b312-45f8-a4d1-00b37f0e57c7%40sessionmgr103)

[Marcus, M., Haden, C.A., & Uttal, D.H. (2017) STEM Learning and
transfer in a children's museum and beyond. *Merrill-Palmer Quarterly*,
63(2):
155-180.](https://www.jstor.org/stable/10.13110/merrpalmquar1982.63.2.0155)
