---
title: "Lab for difference in difference designs"
output: word_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
load("../Planner/appData.Rdata")
```

# Case Study

## Museum Science Spatial: A Program for Pre-Schoolers

A STEM Museum in Florida has had a longstanding partnership with two
researchers at a nearby university. Drs. Andrews and Zheng are
interested in how informal experiences in museums may encourage families
to engage with their children in ways that inspire and support STEM
learning at home. This interest has led them to work with the museum
staff to design a series of exhibits that foster interactions (while at
the museum), while also providing materials to take home, with the hope
of continued engagement. Understanding if this program actually improves
these connections is important, given the continued cost of these
take-home materials. This study focuses on a particular exhibit that
focuses on fostering the development of spatial skills through a series
of building activities.


# Plausible Evaluation Design

## Difference in Difference Analysis

Suppose researchers wish to evaluate the effectiveness of a similar
program on parent interaction with their children at home, measured
using a well-established validated scale based on questions answered
through an online survey collection system. The design is to recruit
families as they enter the museum into the study, as them to fill out a
baseline survey, then randomly assign them to attend the additional
exhibits and receive the additional take home materials (treatment
group) or attend the typical exhibits (control group). After 30 days,
all study families take a survey and their answers for the baseline and
follow-up surveys are used to compute 2 outcome scores for each family,
one for baseline and one for followup.

For simplicity, assume perfect compliance with random assignment and a
perfect response rate.

### Question 1

Using the [planner tool](https://steppcenter.shinyapps.io/Planner/),
find the minimum detectable effect for a `Difference in Differences (DiD)` study in which 20 families are
assigned to the treatment group and 20 families are assigned the control
group when power is .8 for a test with a significance level of .05. For
this question, assume 0 correlation between baseline and followup
surveys.

```{r, include=FALSE}
compute_mdes(V = did_fixed_nocovariate_V(nt = 20,
                                         nc = 20,
                                         ar = 0),
             df=srs_fixed_nocovariate_df(nt = 20,
                                         nc = 20),
             beta = 1-.8,
             alpha = .05
)

```

### Question 2

A member of the study team claims that a better study would use a higher
portion of the study families as treatment. Using the [planner
tool](https://steppcenter.shinyapps.io/Planner/), find the minimum
detectable effect for a `Difference in Differences (DiD)` study in which 30 families are assigned to the
treatment group and 10 families are assigned the control group when
power is .8 for a test with a significance level of .05. For this
question, assume 0 correlation between baseline and followup surveys.

Is this design more sensitive?

```{r, include=FALSE}
compute_mdes(V = did_fixed_nocovariate_V(nt = 30,
                                         nc = 10,
                                         ar = 0),
             df = did_fixed_nocovariate_df(nt = 30,
                                           nc = 10),
             beta = 1-.8,
             alpha = .05
)
```

### Question 3

There is some concern that the study may not be as sensitive as the
researchers need it to be. The **first idea** is to use 60 families,
with 30 in treatment and 30 in control, and continue to assume that the
baselines and followups are not correlated. The **second idea** is to
keep the 40 study members (20 in each group), but assume that the
baseline and followups are likely to be correlated at $\rho = 0.5$.

Using the [planner tool](https://steppcenter.shinyapps.io/Planner/),
find the minimum detectable effect for a `Difference in Differences (DiD)` study in which 30 families are
assigned to the treatment group and 30 families are assigned the control
group, assuming no correlation between baseline and followup surveys,
when power is .8 for a test with a significance level of .05. This is
the sensitivity of the **first idea**.

```{r, include=FALSE}
compute_mdes(V = did_fixed_nocovariate_V(nt = 30,
                                         nc = 30,
                                         ar = 0),
             df = did_fixed_nocovariate_df(nt = 30,
                                           nc = 30),
             beta = 1-.8,
             alpha = .05
)
```

Using the [planner tool](https://steppcenter.shinyapps.io/Planner/),
again, find the minimum detectable effect for a `Difference in Differences (DiD)` study in which 20
families are assigned to the treatment group and 20 families are
assigned the control group, assuming the correlation between baseline
and followup is $\rho = 0.5$, when power is .8 for a test with a
significance level of .05. This is the sensitivity of the **second
idea**.

```{r, include=FALSE}
compute_mdes(V = did_fixed_nocovariate_V(nt = 20, 
                                       nc = 20, 
                                       ar = .5),
             df = did_fixed_nocovariate_df(nt = 20, 
                                         nc = 20),
             beta = 1-.8,
             alpha = .05
)
```

Which idea is more sensitive?

### Question 4

Questions 1-3 involve the same sample parameters as the lab for Chapter
7 (questions A-1 through A-3). Compare the results of this lab to your
answers for that first lab.

# References

[Marcus, M., Haden, C.A., & Uttal, D.H. (2017) STEM Learning and
transfer in a children's museum and beyond. *Merrill-Palmer Quarterly*,
63(2):
155-180.](https://www.jstor.org/stable/10.13110/merrpalmquar1982.63.2.0155)
