---
title: "Lab for randomized block control trials"
subtitle: "Fixed Effects"
output: word_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, include=TRUE) 
load("../Planner/appData.Rdata")
```

# Case Study

## **High School After-School Robotics Program**

Dr. Carlos -- a computer scientist -- and his students have been running
an after-school outreach program focused on robotics at a nearby high
school in Atlanta. This high school predominately serves Black students
-- which are underrepresented in computer science -- and the students
come from both middle-class and low-SES families. In order to be
admitted to the program, which is free, students had to apply and be
recommended by a teacher. Once admitted, students in the program meet
once a week for 2 hours for 16 weeks. During the lessons, they learn to
program and build robots to complete tasks and, at the end of the
program, they take part in a mini 'competition'. The research team has
been excited to see how the students have responded to the program,
including how this has affected their interest in computer science in
general, as well as in studying a STEM related field in college.
Teachers have reported that students in the program are more engaged in
their science classes.

| PICO          | Details                                                                        |
|---------------|--------------------------------------------------------------------------------|
| Pop ulation   | Black high school students interested in robotics in an Atlanta public school. |
| Inter vention | After-school robotics program, 16 weeks x 2 hours                              |
| Com parison   | Unclear                                                                        |
| Outcome       | Belonging (in STEM), Science grades in school, interest in applying to college |

# Plausible Evaluation Design

## Random Assignment Within Blocks (Schools): Fixed Effects

Suppose researchers wish to evaluate the effectiveness of a similar
program on student science grades in school. Researchers enroll students
in the study during the the fall term. Students who are enrolled into
the study are randomized within each school into either the treatment
condition (the program) or a control condition (whatever they planned to
do after school). Spring science scores on a state-mandated standardized
test are used as the outcome. A set of schools are recruited for the
study from a variety of districts through a professional development
conference for state principals and district membership is not
considered in the design. For planning purposes, researchers use a
school-level intraclass correlation of .19 based on work by [Spybrook,
Westine and Taylor
(2016)](https://journals.sagepub.com/doi/10.1177/2332858415625975) (see
Table 4).

The analysis will employ a two-level hierarchical linear model (students
nested within schools) with a random intercept to compare the averages
of the treatment and control students.

For simplicity, assume perfect compliance with random assignment and a
perfect response rate.

### Question 1

Using the [planner tool](https://steppcenter.shinyapps.io/Planner/),
find the minimum detectable effect for a randomized block study with
fixed effects that employed 12 schools, in each of which 10 students are
assigned to the treatment group and 20 students are assigned the control
group when power is .8 for a test with a significance level of .05.

```{r}
compute_mdes(V = blocked_fixed_nocovariate_V(m = 12, 
                                             nt = 10,
                                             nc = 20,
                                             icc = .19), 
             df = blocked_fixed_nocovariate_df(m = 12, 
                                             nt = 10,
                                             nc = 20), 
             beta = 1-.8, 
             alpha = .05 
             )
```

### Question 2

A member of the study team claims that a better study would use a
similar portion of students in treatment and control condition. Using
the [planner tool](https://steppcenter.shinyapps.io/Planner/), find the
minimum detectable effect for a randomized block study with fixed
effects that employed 12 schools, in each of which 15 students are
assigned to the treatment group and 15 students are assigned the control
group when power is .8 for a test with a significance level of .05. Is
this design more sensitive?

```{r}
compute_mdes(V = blocked_fixed_nocovariate_V(m = 12, 
                                             nt = 15,
                                             nc = 15,
                                             icc = .19), 
             df = blocked_fixed_nocovariate_df(m = 12, 
                                             nt = 15,
                                             nc = 15), 
             beta = 1-.8, 
             alpha = .05 
             )

```

### Question 3

After signing up the schools, it was apparent that using a 10th grade
math test as a covariate was possible for the analysis. Earlier analyses
of other cohorts show that the 10th grade math test was correlated with
the focal science test at $\rho = .8$ for an R-square value of
$\rho^2 = .64$. The hope is that having this covariate can reduce the
total number of schools needed from 12 to 10.

Using the [planner tool](https://steppcenter.shinyapps.io/Planner/),
find the minimum detectable effect for a randomized block study with
fixed effects that employed 10 schools, in each of which 15 students are
assigned to the treatment group and 15 students are assigned the control
group for an analysis that also includes the earlier math score, when
power is .8 for a test with a significance level of .05. Is this design
more sensitive?

```{r}
compute_mdes(V = blocked_fixed_covariate_V(m = 10, 
                                             nt = 15,
                                             nc = 15,
                                             icc = .19,
                                           R2_1 = .64,
                                           q = 1), 
             df = blocked_fixed_covariate_df(m = 10, 
                                             nt = 15,
                                             nc = 15,
                                             q = 1), 
             beta = 1-.8, 
             alpha = .05 
             )
```

# References

[Barker, B. S., & Ansorge, J. (2007). Robotics as means to increase
achievement scores in an informal learning environment. *Journal of
research on technology in education*, *39*(3),
229-243.](http://web.b.ebscohost.com.turing.library.northwestern.edu/ehost/pdfviewer/pdfviewer?vid=1&sid=71d566ac-b312-45f8-a4d1-00b37f0e57c7%40sessionmgr103)

[Spybrook, J., Westine, C. D., & Taylor, J. A. (2016). Design parameters
for impact research in science education: A multistate analysis. *AERA
Open*, *2*(1),
2332858415625975.](https://journals.sagepub.com/doi/10.1177/2332858415625975)
