---
title: "Lab for individually randomized control trials"
output: word_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, include=TRUE)
load("../Planner/appData.Rdata")
```

# Case Study

## Museum Science Spatial: A Program for Pre-Schoolers

A STEM Museum in Florida has had a longstanding partnership with two
researchers at a nearby university. Drs. Andrews and Zheng are
interested in how informal experiences in museums may encourage families
to engage with their children in ways that inspire and support STEM
learning at home. This interest has led them to work with the museum
staff to design a series of exhibits that foster interactions (while at
the museum), while also providing materials to take home, with the hope
of continued engagement. Understanding if this program actually improves
these connections is important, given the continued cost of these
take-home materials. This study focuses on a particular exhibit that
focuses on fostering the development of spatial skills through a series
of building activities.

| PICO         | Details                                                    |
|--------------|------------------------------------------------------------|
| Population   | Children ages 3 -- 5 that visit the STEM Museum in Florida |
| Intervention | Construction exhibit, focuses on spatial skills            |
| Comparison   | Unclear                                                    |
| Outcome      | Unclear                                                    |

# Plausible Evaluation Design

## Individual Random Assignment

Suppose researchers wish to evaluate the effectiveness of a similar
program on parent interaction with their children at home, measured
using a well-established validated scale based on questions answered
through an online survey collection system. The design is to recruit
families as they enter the museum into the study, then randomly assign
them to attend the additional exhibits and receive the additional take
home materials (treatment group) or attend the typical exhibits (control
group). After 30 days, all study families take a survey and their
answers are used to compute an outcome score. These scores are then
analyzed using an independent samples $t$ test to compare the averages
of treatment and control families.

For simplicity, assume perfect compliance with random assignment and a
perfect response rate.

### Question 1

Using the [planner tool](https://steppcenter.shinyapps.io/Planner/),
find the minimum detectable effect for a study in which 20 families are
assigned to the treatment group and 20 families are assigned the control
group when power is .8 for a test with a significance level of .05.

```{r}
compute_mdes(V = srs_fixed_nocovariate_V(20,20),
             df=srs_fixed_nocovariate_df(20,20),
             beta = 1-.8,
             alpha = .05
)

```

### Question 2

A member of the study team claims that a better study would use a higher
portion of the study families as treatment. Using the [planner
tool](https://steppcenter.shinyapps.io/Planner/), find the minimum
detectable effect for a study in which 30 families are assigned to the
treatment group and 10 families are assigned the control group when
power is .8 for a test with a significance level of .05. Is this design
more sensitive?

```{r}
compute_mdes(V = srs_fixed_nocovariate_V(30,10),
             df=srs_fixed_nocovariate_df(30,10),
             beta = 1-.8,
             alpha = .05
)
```

### Question 3

There is some concern that the study may not be as sensitive as the
researchers need it to be. One idea is to have each of the 40 study
members fill out a small questionnaire that would lead to a covariate
correlated with the outcome at $\rho = 0.5$ ($\rho^2 = 0.25$). Another
possible use of resources would be to use 60 families, with 30 in
treatment and 30 in control.

Using the [planner tool](https://steppcenter.shinyapps.io/Planner/),
find the minimum detectable effect for a study in which 30 families are
assigned to the treatment group and 30 families are assigned the control
group, without a covariate is employed, when power is .8 for a test with
a significance level of .05. Is this design more sensitive?

```{r}
compute_mdes(V = srs_fixed_nocovariate_V(nt = 30,nc = 30),
             df=srs_fixed_nocovariate_df(nt = 30,nc = 30),
             beta = 1-.8,
             alpha = .05
)
```

Using the [planner tool](https://steppcenter.shinyapps.io/Planner/),
find the minimum detectable effect for a study in which 20 families are
assigned to the treatment group and 20 families are assigned the control
group, with a covariate is employed where ($\rho^2 = 0.25$), when power
is .8 for a test with a significance level of .05.

```{r}
compute_mdes(V = srs_fixed_covariate_V(nt = 20, nc = 20, R2_1 = .25, q = 1),
             df=srs_fixed_covariate_df(nt = 20, nc = 20, q = 1),
             beta = 1-.8,
             alpha = .05
)
```

Which design is more sensitive?

# Reference

[Marcus, M., Haden, C.A., & Uttal, D.H. (2017) STEM Learning and
transfer in a children's museum and beyond. *Merrill-Palmer Quarterly*,
63(2):
155-180.](https://www.jstor.org/stable/10.13110/merrpalmquar1982.63.2.0155)
