---
title: "Lab for non-equivalent control group cluster matching"
subtitle: "Chapter 12"
output: word_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include=TRUE) 
load("../Planner/appData.Rdata")
library(plotly)
```

# Case Study

## **Elementary School Project-Based Learning Program**

Over the past five years, Dr. Wanda Smith and colleagues have been
working closely with Detroit public schools to develop an innovative,
hands-on science program for elementary school teachers and their
students. During this time, Dr. Smith has worked closely with teachers
in three elementary schools to develop and refine a professional
development program and science units focused on environmental science
and biology with children in grades 3-5. These units are hands-on and
project-based and focus not only on scientific knowledge but also on
centering students as scientists and observers in their communities.
Over time, the team has developed five science units, associated
activities, and teaching guidelines. Teachers are deeply engaged in this
work and observations in the classrooms suggest that students are,
indeed, making connections between science and their lives. Student
knowledge of science seems to have increased, as does student
motivation.

+-------------------+----------------------------------------------------------------+
| PICO              | Details                                                        |
+===================+================================================================+
| Population        | Elementary school teachers and students in grades 3-5 in       |
|                   | elementary schools in Detroit serving high percents of         |
|                   | low-SES, majority Black students                               |
+-------------------+----------------------------------------------------------------+
| Intervention      | Professional Development program (teachers) + Science units    |
+-------------------+----------------------------------------------------------------+
| Comparison        | In these schools, science is otherwise taught using textbooks  |
+-------------------+----------------------------------------------------------------+
| Outcome           | Motivation, Science test                                       |
+-------------------+----------------------------------------------------------------+

# Plausible Evaluation Design

## Matching Entire Clusters (Schools)

Suppose researchers wish to evaluate the effectiveness of a similar
program on student science motivation. Researchers enrolled schools in
the program during the summer to serve as treatment schools and now need
to examine the rest of schools in the state to find equivalent matches.
School average of a state test scaled to have a mean of 50 and standard
deviation of 10

### Question 1

Using the [planner tool](https://steppcenter.shinyapps.io/Planner/),
find the minimum detectable effect for a cluster randomized study in
which 6 schools are in the treatment group and 6 schools are in the
control group, in each of which 20 students are surveyed for the
computation of the motivation scale, when power is .8 for a test with a
significance level of .05.

```{r}
compute_mdes(V = clustered_random_nocovariate_V(mt = 6,
                                             mc = 6,
                                             n = 20,
                                             icc = .05), 
             df = clustered_random_nocovariate_df(mt = 6,
                                             mc = 6,
                                             n = 20), 
             beta = 1-.8, 
             alpha = .05 
             )
```

### Question 2

The total sample in Question 1 is 240. A member of the study team claims
that a better study would use 5 schools in treatment and 5 schools in
control and 50 students in each school for a total sample of 500,
arguing that doubling the sample would vastly increase the sensitivity.

Using the [planner tool](https://steppcenter.shinyapps.io/Planner/),
find the minimum detectable effect for a cluster randomized study in
which 5 schools are in the treatment group and 5 schools are in the
control group, in each of which 50 students are surveyed for the
computation of the motivation scale, when power is .8 for a test with a
significance level of .05.

Is this design much more sensitive?

```{r}
compute_mdes(V = clustered_random_nocovariate_V(mt = 5,
                                             mc = 5,
                                             n = 50,
                                             icc = .05), 
             df = clustered_random_nocovariate_df(mt = 5,
                                             mc = 5,
                                             n = 50), 
             beta = 1-.8, 
             alpha = .05 
             )

```

### Question 3

Another researcher claims that instead of doubling the sample, the same
resources can be used to pre-test the students in each school. Earlier
analyses of the motivation instrument show that correlation between
surveys is about $\rho = .9$ at the student level (for an R-square value
of $\rho^2 = .81$) and about $\rho = .3$ at the school level (for an
R-square value of $\rho^2 = .09$).

Using the [planner tool](https://steppcenter.shinyapps.io/Planner/),
find the minimum detectable effect for a cluster randomized study in
which 6 schools are in the treatment group and 6 schools are in the
control group, in each of which 30 students are surveyed for the
computation of the motivation scale, when power is .8 for a test with a
significance level of .05.

Assuming that the strategy for Question 2 is costs the same amount as
this strategy, which produces the more sensitive study?

```{r}
compute_mdes(V = clustered_random_covariate_V(mt = 6,
                                             mc = 6,
                                             n = 20,
                                             icc = .05,
                                             R2_1 = .81,
                                             R2_2 = .09), 
             df = clustered_random_covariate_df(mt = 6,
                                             mc = 6,
                                             n = 20,
                                             q = 1), 
             beta = 1-.8, 
             alpha = .05 
             )
```

### Question 4

Another researcher claims that instead of using a pretest, the same
resources can be used to gather data in more comparison schools. Doing
so may increase the ICC to .1, but could yield 12 comparison schools in
total.

Using the [planner tool](https://steppcenter.shinyapps.io/Planner/),
find the minimum detectable effect for a cluster randomized study in
which 6 schools are in the treatment group and 12 schools are in the
control group, in each of which 20 students are surveyed for the
computation of the motivation scale, but with an ICC of .1, when power
is .8 for a test with a significance level of .05.

```{r}
compute_mdes(V = clustered_random_nocovariate_V(mt = 6,
                                             mc = 12,
                                             n = 20,
                                             icc = .1), 
             df = clustered_random_nocovariate_df(mt = 6,
                                             mc = 12,
                                             n = 20), 
             beta = 1-.8, 
             alpha = .05 
             )
```

Do you think this is a good idea?

# References

[Barron,B. J., Schwartz, D. L., Vye, N. J., Moore, A., Petrosino, A.,
Zech, L., & Bransford, J. D. (1998). Doing with understanding: Lessons
from research on problem-and project-based learning. *Journal of the
learning sciences*, *7*(3-4),
271-311.](https://www-jstor-org.turing.library.northwestern.edu/stable/pdf/1466789.pdf)
